Temporal graph realization from fastest paths

### 2024-01-12 SAND2024/SAND24-longversion.tex
before new changes

From the construction of $Z$ (\ie by exhaustively removing vertices of degree one from~$G$), it follows that $Z$ consists of disjoint trees. 
Each of these trees has a unique neighbor $u_i$ in $G'$. Denote by $T_i$ the tree that is obtained by such a vertex $u_i$ of $G'$ and all the trees of $Z$ that are incident to $u_i$. Then  we call $u_i$ the \emph{clip vertex} of the tree $T_i$. 
For each of these trees~$T_{i}$, we select an arbitrary vertex $r_i \in Z$ that is a neighbor of the clip vertex $u_i$ in $T_i$,
and call it the \emph{representative vertex} of the tree $T_{u_i}$. 
We now define by $Z^*$ the set of all representatives $r_i$ of the trees $T_i$,
where the clip vertex $u_i$ of $T_i$ is a vertex of interest, \ie
\[
Z^* = \{ r_i  :  r_i \in Z \text{ is a neighbor of the clip vertex $u_i$ in the tree $T_i$, and $u_i \in U$}\}.
\]
Since there are $O(k)$ vertices of interest, we get that $|Z^*| \in O(k)$.
Finally, the set of \emph{important vertices} is defined as the set $U \cup U^{\ast} \cup Z^{\ast}$.
For an illustration see \cref{fig:labelingVertices}. Note that determining sets $U$, $U^*$, and $Z^*$ takes linear time.

@@

In the case when segment has $4$ edges, it can happen that the fastest temporal paths from above, do not traverse $f$. But in this case we know that three edges have the same label, denote it $c$, while the label of the last one is a function of $c$.
In the case when the segment $S_{u,v}$ has just three, two or one edge, this procedure does not improve anything,
therefore these segments remain with three, two or one not yet determined edges, respectively.

@@

At this point $G'$ is a graph, where each edge $e$ has a value for its label $\lambda(e)$
that depends on (\ie is a function of) some other label $\lambda(f)$ of edge $f$,
or it depends on no other label.
We now describe how we create variables and start building our ILP instances. 
For every edge $e$ in $G'$ that is incident to a vertex of interest we create a variable $x_e$ that can have values from $ \{1, 2, \dots, \Delta\}$.
Besides that we create one variable for each edge that is still not yet determined on a segment.
Since each vertex of interest is incident to at most $k$ edges, and each segment has at most one extra not yet determined edge, we create $O(k^2)$ variables.
At the end we create our final guess.
\begin{enumerate}[G-1.]
% counter for guesses continues
\setcounter{enumi}{\value{guesscounter}}
%%
    \item \label{FPT:guessallPermutations}
We guess the permutation of all $O(k^2)$ variables,
together with the relation of each variable to the labels of edges incident to these not yet determined edges.
Namely, for an edge $e$ that is not yet determined, we set its value to $x_e$ and check labels of all of its neighbors,
which are determined by some other label,
and variables of the not yet determined neighbors,
and guess if $x_e$ is smaller, equal or bigger than the labels of the edges of its neighbors.
So, for any two variables $x_e$ and $x_f$, we know if $x_e < x_f$ or $x_e = x_f$, or $x_e > x_f$,
and for any neighboring edge $g$ of $e$ we know if $x_e < \lambda(g)$ or 
$x_e = \lambda(g)$, or $x_e > \lambda(g)$.
This results in $O(k^2)! = k^{O(k^2)}$ guesses
and consequently
each of the ILP instances we created up to now is further split into $k^{O(k^2)}$ new ones.
\end{enumerate}
We have now finished creating all ILP instances.
From \cref{sec:FPT-guessing} we know the structure of all guessed paths, to which we have just added also the knowledge of permutation of all variables.
We proceed with adding constraints to each of our ILP instances.
First we add all constraints for the labels of edges that we have determined up to now.
We then continue to iterate through all pairs of vertices and start adding equality (resp.~inequality) constraints for the fastest (resp.~not necessarily fastest) temporal paths between them.

@@

\subparagraph{\boldmath Considering the paths for vertices from $Z$.}
All of the above is enough to determine the labeling $\lambda$ of $G'$. Now we have to make sure that the labeling considers also the vertices of~$Z= V(G) \setminus V(G')$ that we initially removed from $G$.
Recall that $Z$ consists of disjoint trees, and that each of these trees has a unique neighbor $u_i$ in $G'$. 
Furthermore recall that $T_i$ denotes the tree that is obtained by such a vertex $u_i$ of $G'$ and all the trees of $Z$ that are incident to $u_i$; then $u_i$ is the \emph{clip vertex} of the tree $T_i$. 
Moreover, for each of these trees~$T_{i}$ we have selected an arbitrary vertex $r_i \in Z$ that is a neighbor of the clip vertex $u_i$ in $T_i$, 
and $r_i$ is called the \emph{representative vertex} of the tree $T_{u_i}$. 
Let us now describe now how to determine the labels on the edges of any such tree $T_i$.
To determine the correct label of all edges of $T_i$ we use the following property.
\begin{lemma}
Let $T_i$ be a tree having $u_i$ as its clip vertex, and let $e_i = (u_i,r_i)$. 
Among all neighbors of $u_i$ in $G'$, let $v \in N_{G'}(u_i)$ be the one that is closest vertex to $r_i$ (with respect to the values of $D$), \ie  $D_{r_i,v} \leq D_{r_i,w}$ for all $w \in N_{G'}(u_i)$.
Then the path $P^*=(r_i,u_i,v)$ is the fastest temporal path from $r_i$ to $v$.
\end{lemma}
\begin{proof}
Suppose that this is not true.
Then there exists a faster path $P^*_2$ from $r_i$ to $v$, that goes through the clip vertex $u_i$ of $T_i$ (as this is the only neighbor of $r_i$), through some other vertex $w \in N_{G'}(u) \setminus \{v \}$, 
and through some other path $P'$ in $G$, before it finishes in $v$, where $P'$ is at least an edge (from $w$ to $v$). 
Therefore $P^*_2= (r_i,u_i,w,P',v)$, where $d(P^*_2) \leq d(P^*)$.
Now since $D_{r_i,v} \leq D_{r_i,w}$ for all $ w\in N_{G'}(u)$ the first part of path $P^*_2$ from $r_i$ to $w$ takes at least $D_{r_i,v}$ time.
Since $v \neq w$ we need at least one more time-step (one more edge) to traverse from $w$ to reach $v$. 
So $d(P^*_2) \geq D_{r_i,v} + 1$, and so $P^*_2$ cannot be faster than $P^*$.
\end{proof}
Suppose now that we know that 
$(r_i,u_i,v)$ is the fastest temporal path from the representative $r_i$ of $T_i$ to the vertex $v$ in the neighborhood of the clip vertex $v_i$ of $T_i$.
Then
we can determine the label of edge $r_i u_i$ as 
$\lambda (r_i u_i) \equiv \lambda(u_i v) + 1 - D_{r_i,v} \pmod \Delta$.
Now, using the algorithm for trees (see \cref{thm:deltaExact-PolyTimeTrees}), we determine labels on all edges of $T_i$.
We repeat this procedure for all trees in $Z$.
What remains, is to add the equality (resp.~inequality) constraints for the fastest (resp. non-fastest) temporal paths from vertices of $Z$ to all other vertices in $G'$ and vice versa.
Note that since 
there is a unique path between vertices of tree $T_i$, and since all edges of tree $T_i$ are determined with respect to the same label,
we present the study for cases when we find fastest temporal paths from and to the 
representative vertex $r_i$ of tree $T_i$.
Each of these paths are then uniquely extended to all vertices in $T_i$.

>>>>

From the construction of $Z$ (\ie by exhaustively removing vertices of degree one from~$G$), it follows that $G[Z]$ (graph induced in $G$ by $Z$) consists of disjoint trees. 
Each of these trees has a unique neighbor $u_i$ in $G'$. Denote by $T_i$ the tree that is obtained by such a vertex $u_i$ and all the trees from $G[Z]$ that are incident to $u_i$. Then  we call $u_i$ the \emph{clip vertex} of the tree $T_i$. 
For each of these trees~$T_{i}$, we select an arbitrary vertex $r_i \in Z$ that is a neighbor of the clip vertex $u_i$ in $T_i$,
and call it the \emph{representative vertex} of the tree $T_{u_i}$. 
We now define by $Z^*$ the set of all representatives $r_i$ of the trees $T_i$,
where the clip vertex $u_i$ of $T_i$ is a vertex of interest, \ie
\[
Z^* = \{ r_i  :  r_i \in Z \text{ is a neighbor of the clip vertex $u_i$ in the tree $T_i$, and $u_i \in U$}\}.
\]
Since there are $O(k)$ vertices of interest, we get that $|Z^*| \in O(k)$.
Finally, the set of \emph{important vertices} is defined as the set $U \cup U^{\ast} \cup Z^{\ast}$.
For an illustration see \cref{fig:labelingVertices}. Note that determining sets $U$, $U^*$, and $Z^*$ takes linear time.

@@

In the case when a segment has $4$ edges, it can happen that the fastest temporal paths from above, do not traverse $f$. But in this case, we know that three edges have the same label, denote it $c$, while the label of the last one is a function of $c$.
In the case when the segment $S_{u,v}$ has just three, two, or one edge, this procedure does not improve anything,
therefore these segments remain with three, two, or one not yet determined edges, respectively.

@@

At this point $G'$ is a graph, where each edge $e$ has a value for its label $\lambda(e)$
that depends on (\ie is a function of) some other label $\lambda(f)$ of edge $f$,
or it depends on no other label.
We now describe how we create variables and start building our ILP instances. 
For every edge $e$ in $G'$ that is incident to a vertex of interest, we create a variable $x_e$ that can have values from $ \{1, 2, \dots, \Delta\}$.
Besides that, we create one variable for each edge that is still not yet determined on a segment.
Since each vertex of interest is incident to at most $k$ edges, and each segment has at most one extra not yet determined edge, we create $O(k^2)$ variables.
At the end we create our final guess.
\begin{enumerate}[G-1.]
% counter for guesses continues
\setcounter{enumi}{\value{guesscounter}}
%%
    \item \label{FPT:guessallPermutations}
We guess the permutation of all $O(k^2)$ variables,
together with the relation of each variable to the labels of edges incident to these not yet determined edges.
Namely, for an edge $e$ that is not yet determined, we set its value to $x_e$ and check the labels of all of its neighbors,
which are determined by some other label,
and variables of the not yet determined neighbors,
and guess if $x_e$ is smaller, equal or bigger than the labels of the edges of its neighbors.
So, for any two variables $x_e$ and $x_f$, we know if $x_e < x_f$ or $x_e = x_f$, or $x_e > x_f$,
and for any neighboring edge $g$ of $e$ we know if $x_e < \lambda(g)$ or 
$x_e = \lambda(g)$, or $x_e > \lambda(g)$.
This results in $O(k^2)! = k^{O(k^2)}$ guesses
and consequently
each of the ILP instances we created up to now is further split into $k^{O(k^2)}$ new ones.
\end{enumerate}
We have now finished creating all ILP instances.
From \cref{sec:FPT-guessing} we know the structure of all guessed paths, to which we have just added also the knowledge of permutation of all variables.
We proceed with adding constraints to each of our ILP instances.
First we add all constraints for the labels of edges that we have determined up to now.
We then continue to iterate through all pairs of vertices and start adding equality (resp.~inequality) constraints for the fastest (resp.~not necessarily fastest) temporal paths between them.

@@

\subparagraph{\boldmath Considering the paths for vertices from $Z$.}
All of the above is enough to determine the labeling $\lambda$ of $G'$. Now we have to make sure that the labeling considers also the vertices of~$Z= V(G) \setminus V(G')$ that we initially removed from $G$.
Recall that $Z$ consists of disjoint trees, and that each of these trees has a unique neighbor $u_i$ in $G'$. 
Furthermore, recall that $T_i$ denotes the tree that is obtained by such a vertex $u_i$ of $G'$ and all the trees of $Z$ that are incident to $u_i$; then $u_i$ is the \emph{clip vertex} of the tree $T_i$. 
Moreover, for each of these trees~$T_{i}$ we have selected an arbitrary vertex $r_i \in Z$ that is a neighbor of the clip vertex $u_i$ in $T_i$, 
and $r_i$ is called the \emph{representative vertex} of the tree $T_{u_i}$. 
Let us now describe now how to determine the labels on the edges of any such tree $T_i$.
To determine the correct label of all edges of $T_i$ we use the following property.
\begin{lemma}
Let $T_i$ be a tree having $u_i$ as its clip vertex, and let $e_i = (u_i,r_i)$. 
Among all neighbors of $u_i$ in $G'$, let $v \in N_{G'}(u_i)$ be the one that is closest vertex to $r_i$ (with respect to the values of $D$), \ie  $D_{r_i,v} \leq D_{r_i,w}$ for all $w \in N_{G'}(u_i)$.
Then the path $P^*=(r_i,u_i,v)$ is the fastest temporal path from $r_i$ to $v$.
\end{lemma}
\begin{proof}
Suppose that this is not true.
Then there exists a faster path $P^*_2$ from $r_i$ to $v$, that goes through the clip vertex $u_i$ of $T_i$ (as this is the only neighbor of $r_i$), through some other vertex $w \in N_{G'}(u) \setminus \{v \}$, 
and through some other path $P'$ in $G$, before it finishes in $v$, where $P'$ is at least an edge (from $w$ to $v$). 
Therefore $P^*_2= (r_i,u_i,w,P',v)$, where $d(P^*_2) \leq d(P^*)$.
Now since $D_{r_i,v} \leq D_{r_i,w}$ for all $ w\in N_{G'}(u)$ the first part of path $P^*_2$ from $r_i$ to $w$ takes at least $D_{r_i,v}$ time.
Since $v \neq w$ we need at least one more time-step (one more edge) to traverse from $w$ to reach $v$. 
So $d(P^*_2) \geq D_{r_i,v} + 1$, and so $P^*_2$ cannot be faster than $P^*$.
\end{proof}
Suppose now that we know that 
$(r_i,u_i,v)$ is the fastest temporal path from the representative $r_i$ of $T_i$ to the vertex $v$ in the neighborhood of the clip vertex $v_i$ of $T_i$.
Then
we can determine the label of edge $r_i u_i$ as 
$\lambda (r_i u_i) \equiv \lambda(u_i v) + 1 - D_{r_i,v} \pmod \Delta$.
Now, using the algorithm for trees (see \cref{thm:deltaExact-PolyTimeTrees}), we determine labels on all edges of $T_i$.
We repeat this procedure for all trees in $Z$.
What remains, is to add the equality (resp.~inequality) constraints for the fastest (resp. non-fastest) temporal paths from vertices of $Z$ to all other vertices in $G'$ and vice versa.
Note that since 
there is a unique path between vertices of tree $T_i$, and since all edges of tree $T_i$ are determined with respect to the same label,
we present the study for cases when we find the fastest temporal paths from and to the 
representative vertex $r_i$ of tree $T_i$.
Each of these paths is then uniquely extended to all vertices in $T_i$.
